BOLD Parameter Estimation Using Particle Filters

### 2010-07-20 doc/thesis/intro.tex
additions to intro

\chapter{Introduction}
For the past twenty years Functional Magnetic Resonance Imaging (FMRI) 
has been at the forefront of cognitive research. Despite it's
limited temporal resolution, FMRI is the standard tool for localizing 
neural activation.  Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis relatively straight forward.

By modeling the governing equations behind the neural response that
drives FMRI, it is possible to make the analyses of FMRI even more
powerful. The underlying state equations hold important information
about how individual brain regions react to stimuli. The model parameters
on the other hand, hold important information about the patients individual
physiology including existing and future pathologies. In short,
the long chain of events driving observable FMRI signal contains more
information than which regions respond to which stimuli.

In the past fifteen years, a steady stream of studies have built
on the original Blood Oxygen Level Dependent (BOLD) signal 
derivation first described by \cite{Ogawa}.
The seminal work by \cite{Buxton1998} attempted to explain the
time evolution of the BOLD signal using a windkessel model to
describe the time local changes in Deoxygenated Hemoglobin content.
Various papers made substantial improvements to this model until
\cite{Buxton2004} brought all the changes together into a single complete 
set of equations. And while there have been numerous adaptations in the model, 
many of them summarized in \cite{Deneux2006}, even the most basic version
has been shown to have less bias error than the convolution based 
"Canonical Hemodynamic Model" \cite{Deneux2006}, \cite{Handwerker2004}.
On the other hand many of the BOLD signal models have far more parameters
than a simple scaling parameter. In fact, the number of parameters
range from seven \cite{Riera2004} to 50 \cite{Behzadi2005} per
signal time course; a signal which may be as short as 100 samples long.
Thus, even in a small FMRI image of the brain (20x20x20), the number of parameters 
could easily exceed 10,000. Clearly this number of parameters presents
a significant risk of being under-determined and could suffer catastrophic
variance error; to make no mention of the computation cost. 
In this work a method of countering these problem is presented
with the use of a particle filter.

>>>>

\chapter{Introduction}
Traditional methods of analyzing images produced by Function Magnetic Resonance 
Imaging (FMRI) use regression techniques with very few degrees of freedom. 
Though adding more degrees of freedom naturally mandates more computation,
in this thesis I will discuss a Sequential Monte Carlo method of fitting a nonlinear
model with seven degrees of freedom at 
a computation cost that would allow real time calculations for a multiple voxels
(4 dimensional pixels). More practically, the method described here forms a completely
separate method of detecting neural activity in comparison to the traditionally
used Statistical Parametric Mapping (SPM). Though more computationally intense,
this method is capable of modeling nonlinear effects, is conceptually simpler
and provides more detailed output. Additionally,
by using a separate particle filter for each voxel's time
series it should be feasible to calculate parameters and make real-time predictions
for small neural regions, a feature which could be useful towards real time FMRI 
(\cite{DeCharms2005}).

For the past twenty years Functional Magnetic Resonance Imaging (FMRI) 
has been at the forefront of cognitive research. Despite it's
limited temporal resolution, FMRI is the standard tool for localizing 
neural activation.  Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is quick and cheap, and its analysis straight forward.
By modeling the governing equations behind the neural response that
drives FMRI, it is possible to increase the power of FMRI.
The underlying state equations hold important information
about how individual brain regions react to stimuli. The model parameters
on the other hand, hold important information about the patients individual
physiology including existing and future pathologies. In short,
the long chain of events driving FMRI signals contain information 
beyond correlation with stimuli.

In the past fifteen years, a steady stream of studies have built
on the original Blood Oxygen Level Dependent (BOLD) signal 
derivation first described by \cite{Ogawa}.
The seminal work by \cite{Buxton1998} attempted to explain the
time evolution of the BOLD signal using a windkessel model to
describe the local changes in Deoxygenated Hemoglobin content.
Various papers made substantial improvements to this model until
\cite{Buxton2004} brought all the changes together into a single complete 
set of equations. And while there have been numerous adaptations in the model, 
many of them summarized in \cite{Deneux2006}, even the basic versions
have less bias error than the empirically driven \emph{Canonical Hemodynamic Model}
\cite{Deneux2006}, \cite{Handwerker2004}.
On the other hand many of the BOLD signal models have numbers
of parameters ranging from seven \cite{Riera2004} to 50 \cite{Behzadi2005} 
for a signal as short as 100 samples long. Clearly this number of parameters presents
a significant risk of being under-determined and having high computation cost. 
In this work, only the simplest physiologically inspired model will be
used (with 7 parameters), and steps will be taken to make the most of computation
time.
